---
title: "R Notebook following ModernStatisticswithR.com"
author: Caroline Sanicola
date: June 5, 2024
output: 
  html_notebook:
  toc: true
  toc_float: true
  toc_depth: 2
  number_sections: true
  theme: slate
  highlight: default
editor_options: 
  markdown: 
    wrap: 72
---

# 2. The Basics
- for R, the best application to use is RStudio which you need to download in addition to the language R
- when using R notebooks, you can insert code chunks (use `ctrl + alt + i` to insert for windows)

```{r}
# scripts can use either just one line of code
1+1
# or multiple lines of code
2*2
1+2*3-5
```
- the main two concepts for R are **variables** (which are used for storing data) and **functions** (used for doing things with the data)
```{r}
# to store data in R you do the below to a variable name
x <- 4
# (shortcut for the `<-` is `alt + -`)
```
- once data is stored in a variable name, then you can do things with that variable/data
```{r}
x + x
```
- R is case sensitive for storing variables so the below are all different variables that can be used to store data:
```{r}
my_variable <- 5
MY_VARIABLE <- 6
My_Variable <- 7
```

- variables aren't stored if there are spaces in the name or special characters 

- to turn multiple lines into comments, you can use the shortcut: `Ctrl + Shift + C`

## Exercise
1. What happens if you use an invalid character in a variable name?

net income <- income - taxes
net-income <- income - taxes
ca$h <- income - taxes

these will all just throw errors and won't let you save the variables 

2. What happens if you put R code as a comment? For example,
```{r}
income <- 100
taxes <- 20
net_income <- income - taxes
# gross_income <- net_income + taxes
```
the variable commented out will not show up in the environment because it isn't being saved as a useable variable

3. What happens if you remove a line break and replace it by a semicolon ;? For example,
```{r}
income <- 200; taxes <- 30
```
these will save as two separate variables -> the semicolon acts as a line separator 

4. What happens if you do two assignments on the same line? For example,
```{r}
income2 <- taxes2 <- 100
```
this saves both variables with the same number 

























# 3. The cornerstones of statistics


# 4. Exploratory Data Analysis and Unsupervised Learning

# 5. Dealing with messy data 

# 6. R programming

# 7. The role of simulation in modern statistics

# 8. Regression models 

# 9. Survival analysis and censored data 

# 10. Structural equation models

# 11. Predictive modeling and machine learning

# 12. Advanced topics 

# 13. Debugging

# 14. Mathematical appendix

