---
title: "multiple_regression_in_R"
output: html_document
---

# StatQuest: Doing Multiple Regression in R
[Source](https://www.youtube.com/watch?v=hokALdIst8k)

# Loading Data 
```{r}
# first create the dataframe which we will be working with 
mouse.data <- data.frame(
  weight=c(0.9, 1.8, 2.4, 3.5, 3.9, 4.4, 5.1, 5.6, 6.3),
  size=c(1.4, 2.6, 1.0, 3.7, 5.5, 3.2, 3.0, 4.9, 6.3),
  tail=c(0.7, 1.3, 0.7, 2.0, 3.6, 3.0, 2.9, 3.9, 4.0)
)
# you can also just load in an existing dataframe too
mouse.data
```

- for our simple regression, our question will be: how well does weight predict size?
# Simple Regression
# Step 1: Plot the data
```{r}
plot(mouse.data$weight, mouse.data$size)
# this first step is an important part in seeing if there even is a relationship in the data
```

# Step 2: Use the lm() (linear model) function to fit a line to the data
```{r}
simple.regression <- lm(size ~ weight, data=mouse.data)
# in R, we specify the relationship we are looking for by using the `~` character
# in this example, we specify size is predicted by weight 
summary(simple.regression)
```
- from the summary table, we are specifically interested in the R² and adjusted R²: `Multiple R-squared:  0.6133,	Adjusted R-squared:  0.558`

# Step 3: Plot the new regression line to the previous plot:
```{r}
plot(mouse.data$weight, mouse.data$size)
abline(simple.regression, col="red", lwd=2)
```

# Multiple Regression
- for this, our question will be does weight and tail predict size?

# Step 1: Plot the data 
```{r}
plot(mouse.data)
# since no x and y-axes were specified, R plots all the combinations against one another
# based on the weight and tail plots, they look correlated which means they provide similar information and that we might not need both in our model
```

# Step 2: Use the lm() (linear model) function to fit a plane to the data 
```{r}
multiple.regression <- lm(size ~ weight + tail, data=mouse.data)
# the actual formula being used is: `size = y-intercept + slope1 x weight + slope2 x tail`
# we are specifying that size is predicted by weight and tail 
summary(multiple.regression)
```

weight       -0.3293     0.3933  -0.837   0.4345 
>this line compares the multiple regression to the simple regression
the p-value here means that using weight and tail isn't significantly better than using tail alone to predict size 

tail          1.6470     0.5363   3.071   0.0219 *
>this line also is comparing the multiple to simple regression 
the p-value here means that using weight and tail is significantly better than using weight alone to predict size 

**In Summary**
Using weight and tail to predict size is good but if we want to save time, we could spare ourselves from collecting weight data and just use tail lengths to predict size 








































