---
title: "Sr Data Analysis Project1"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# Analyzing Healthcare Data to see about Practice Trends over a Period of Time

# Loading in Packages and Data

```{r}
install.packages('tidyverse')
install.packages('GGally')
install.packages('gt')
install.packages('gtsummary')
install.packages('ggplot2')

library(tidyverse)
library(HDSinRdata)
library(GGally)
library(gt)
library(gtsummary)
library(ggplot2)
```

First we need to load in the data
This data is from the [Sr_Data_Analyst_Project1 Notebook]('https://github.com/csanicola/important-reference-repo/blob/822febba0fca0a5ba7bc20c3fe333594ffe51ce6/Practice/sr_data_analyst_project1.ipynb') that created fake healthcare data for analysis purposes

We then want to look at the head of the data to see what it looks like and if the correct csv was loaded in.
```{r}
df <- read.csv('D:/GitHub/important-reference-repo/Data/fake_healthcare_data_v3.csv')
head(df)
```


# Handle NA/Missing Data
```{r}
sum(is.na(df))
```


# Overview of the Data

To get a bit more detailed overview of what type of data is in this dataframe, we can use the `summary()` function
```{r}
summary(df)
```
First need to convert the categorical values as factors, to make everything easier, especially plotting.
```{r}
df$Practice_ID <- as.factor(df$Practice_ID)
df$Quarter <- as.factor(df$Quarter)
df$Year <- as.factor(df$Year)
```


```{r}
# basic boxplot
plot(df$Practice_ID, df$Visit_Volume, ylab="Visit Volume", xlab="Practice")
```
```{r}
ggplot(df, aes(x = Year, y = Visit_Volume)) +
  geom_bar(stat="summary", fun="mean") +
  labs(y="Visit Volume", x="Practice")
```
We can see two categorical values like the year and quarter in a boxplot
```{r}
boxplot(Avg_Wait_Time_Min ~ Year + Quarter, data=df,
        col=c("orange", "blue", "red", "yellow", "pink"),
        xlab="Year + Quarter", ylab="Average Wait Time")
```

If we now want to see the correlation of two continuous variables (Avg_Wait_Time_Min and Patient_Satisfaction_Score), we would use a scatterplot
```{r}
plot(df$Avg_Wait_Time_Min, df$Provider_Productivity, col="blue",
     xlab="Average Wait Time (min)",
     ylab="Patient Satisfaction Score")
```
There doesn't look to be any correlation, but we can check by running a Pearsons and Spearman correlation calculation
```{r}
cor(df$Avg_Wait_Time_Min, df$Provider_Productivity)
cor(df$Avg_Wait_Time_Min, df$Provider_Productivity, method="spearman")
```

If we want to see all of the univariate and bivariate plot combinations, we can use the `ggpairs()` function from the GGally package
```{r}
ggpairs(df)
```
```{r}
ggcorr(df, label=TRUE)
```
```{r}
tbl_summary(df)
```

We can also summarize the data by a specific variable, so if we use Year, it would look like this:
```{r}
tbl_summary(df, by="Year",
            statistic=list(all_continuous() ~ "{mean} ({sd})")) |> 
  as_gt()
```














